Interspecies comparisons of Mg/Ca ratios in limpet shells

### 2024-06-24 Manuscript.qmd
added separate graphs

@@

```{r SST Data}

@@

```{r SST plot}
#| fig-cap: SST values averaged from  years. SST values were sourced from @Good2020-nl
#| fig-height: 5
#| fig-width: 5
#| label: fig-SSTs

@@

@@

@fig-Nac_iso and @fig-Pat_iso show the previously acquired sequential δ^18^O-values of limpet shells by species. Due to the high sampling resolution, all shells show seasonally resolved and multi-year quasi-sinusoidal records of SST change , with *Nacella* sp. shells living 2–3 years and *Patella v*. regularly over 3 years. In those instances where longer lifetimes are encountered (i.e. specimen QG1-7246-1, @fig-Pat_iso b). Interestingly, the ranges in δ^18^O-values are not similar between the coeval *Nacella* sp. specimens with δ^18^O values for the previous summer temperatures ranging from 0.4‰ to above 1.0‰, and for winter values between 3.0‰ and 2.0‰. *P. vulgata* has similarly variable δ^18^O-values for summer and winter seasons, alas the lack of more that one modern specimens, means that we cannot quantify the variability of recorded annual ranges between specimens.

```{r Nacella_Isotope_plot}
#| fig-cap: Isotope values of Nacella sp. limpet shells from Cambaceres Bay (*a-f*).
#| fig-height: 10
#| fig-width: 10
#| label: fig-Nac_iso
# Filter data for Modern period
Nacella_iso <- D18O %>% filter(grepl("Nacella", species))

@@

```{r Patella_Isotope_plot}
#| fig-cap: Isotope values of Patella vulagata limpet shells from Quoygrew (QG, *a-g*) and Rack Wick Bay (ORK, *h*) on Westray, Orkney.
#| fig-height: 10
#| fig-width: 10
#| label: fig-Pat_iso
# Filter data for Modern period
Patella_iso <- D18O %>% filter(grepl("Patella", species))

@@

## *Nacella deaureata*

## *Nacella magellanica*

@@

#

```{r Correlation Graphs }
#| fig-cap: Correlation graphs for all specimens
#| fig-height: 20
#| fig-width: 10

@@

@@

@@

# References
```{r Modern iso ranges}
# Ranges <- D18O %>%
#   filter(!(sampleID == "ORK_LT5" & d18O < 0.2)) %>% #excluding outlier
# group_by(sampleID,period) %>%
#   mutate(max=max(d18O),
#          min=min(d18O)) %>%
#   select(min, sampleID,max) %>% unique()
# 
# Modern_Ranges <- Ranges %>%
#   filter(period=="Modern") %>%
#   arrange(range)

@@

>>>>

D18O <-   D18O %>% filter(sampleID %in% MgCa$sampleID)

@@

```{r SST_Data}

@@

```{r SST_plot}
#| fig-cap: SST values averaged from  years. SST values were sourced from @Good2020-nl
#| fig-height: 5
#| fig-width: 5
#| label: fig-SSTs

@@

Using this system we carried out elemental imaging of the shell specimens at a resolution of 50 µm distance between sampling spots. Each spot was irradiated 10 times with the 3 first spectra used for cleaning steps and the remaining summed to get an average Mg/Ca intensity ratio for each sample spot. Subsequently, we re-sampled the section using a line scan at 10 µm resolution. This leads to an overlap between sample spots, but also allows for a continuous record without gaps. Intensity ratios were filtered for cases with high relative standard deviation (i.e. more than 10%). These occurred in places where the previous sampling procedures for carbonate powder as part of the oxygen isotope analysis left an uneven sample surfaces introducing variability in the plasma generation and thus uncertainty in the data of these locations.

@@

@fig-Nac_iso and @fig-Pat_iso show the previously acquired sequential δ^18^O-values of limpet shells by species. Due to the high sampling resolution, all shells show seasonally resolved and multi-year quasi-sinusoidal records of SST change , with *Nacella* sp. shells living 2–3 years and *Patella v*. regularly over 3 years. In those instances where longer lifetimes are encountered (i.e. specimen QG1-7246-1, @fig-Pat_iso b). Interestingly, the annual ranges in δ^18^O-values are not similar between the coeval *Nacella* sp. specimens with δ^18^O values for the previous summer temperatures ranging from 0.4‰ to above 1.0‰, and for winter values between 2.7‰ and 2.0‰. *P. vulgata* has similarly variable δ^18^O-values for summer and winter seasons, alas the lack of more that one modern specimens, means that we cannot quantify the variability of recorded annual ranges between specimens.

```{r Nacella_Isotope_plot}
#| fig-cap: Isotope values of *Nacella* sp. limpet shells from Cambaceres Bay (*a-f*).
#| fig-height: 10
#| fig-width: 10
#| label: fig-Nac_iso
# Filter data for Modern period
Nacella_iso <- D18O %>% filter(grepl("Nacella", species))

@@

```{r Patella_Isotope_plot}
#| fig-cap: Isotope values of *Patella vulgata* limpet shells from Quoygrew (QG, *a-g*) and Rack Wick Bay (ORK, *h*) on Westray, Orkney.
#| fig-height: 10
#| fig-width: 10
#| label: fig-Pat_iso
# Filter data for Modern period
Patella_iso <- D18O %>% filter(grepl("Patella", species))

@@

## Dynamic Time Warping

Because the exact location used in the analysis of stable oxygen isotopes is not possible to reconstruct for each carbonate powder sample, we used dynamic time warping (DTW) to align the time series of δ^18^O-values and Mg/Ca ratios. DTW is an algorithm that measures similarity between two proxy-sequences, which may vary in sampling resolution or interval. By stretching or compressing sections of the series, DTW finds the probable alignment between the two sequences. This allows us to compare the proxy data sets more effectively, ensuring that the temporal dynamics of each shell are accurately matched despite possible discrepancies in sampling intervals or rates. We applied the DTW algorithm using the `dtw` package in R [@Giorgino2009-sj; @R_Core_Team2020-mk], which provides a robust framework for aligning time series data. This involved selecting appropriate distance measures and constraints to ensure meaningful alignment. The process involved iterative adjustments to minimise the overall distance between corresponding points in the two data sets. The code used for this analysis as well as the data required to run it can be found in this **OSF repository**.

@@

-   Show one or a few of the LIBS maps as well as the line scans.
-   Show one the line scans together with isotope data.
-   Show correlation graphs.

```{r Patella_Correlation_Graphs }
#| fig-cap: Correlation graphs for *Patella vulgata* specimens
#| fig-height: 20
#| fig-width: 15

P_all <- Corr_data %>% 
  pivot_wider(values_from = "value",names_from = "proxy") %>% 
  # mutate(region = substr(sampleID, 1, 1)) %>% 
  left_join(D18O %>% select(sampleID,period,species) %>% unique(),by="sampleID") %>% 
  filter(species=="Patella vulgata") %>% 
  ggplot()+
  aes(d18o,mg_ca,col=sampleID)+
  geom_point(show.legend = FALSE)+
  labs(x=expression("‰ δ" ^ "18" * "O [VPDB]"),y="Mg/Ca intensity ratio")+
  stat_smooth(aes(fill = sampleID,col=sampleID), method = "lm", formula = y~x,show.legend = FALSE) +
  # stat_regline_equation(label.x = 2, #label.y = 32
  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
  #   formula = y~x
  # ) +
  scale_colour_manual(values = c("#542788","#007f7f", "#77b0a3","black", "#1f78b4","grey50", "#d73027", "#EC7063")) +
scale_fill_manual(values = c( "#542788","#007f7f", "#77b0a3","black", "#1f78b4","grey50", "#d73027", "#EC7063"))#+
  # facet_wrap(~region,ncol=2,scales = "free")

@@

P_individual <- P_all+ylim(0,0.8)+
  facet_wrap(~sampleID,ncol=4)+#ylim(0.15,NA)+
  ggpubr::stat_regline_equation(label.x = 0.2, label.y = 0.65,
    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,size=6,
    formula = y~x
  )+
  theme_cowplot(20)+
  theme(
    strip.background = element_rect(fill = "white", colour = NA),
    strip.text = element_text(colour = "black"),
    panel.border = element_rect(colour="black")
  )

@@

## *Nacella* sp.

```{r Nacella_Correlation_Graphs}
#| fig-cap: Correlation graphs for *Nacella* sp. specimens
#| fig-height: 15
#| fig-width: 15

P_all <- Corr_data %>% 
  pivot_wider(values_from = "value",names_from = "proxy") %>% 
  left_join(D18O %>% select(sampleID,period,species) %>% unique(),by="sampleID") %>% 
  filter(grepl("Nacella", species)) %>% 
  ggplot()+
  aes(d18o,mg_ca,col=sampleID)+
  geom_point(show.legend = FALSE)+
  labs(x=expression("‰ δ" ^ "18" * "O [VPDB]"),y="Mg/Ca intensity ratio")+
  stat_smooth(aes(fill = sampleID,col=sampleID), method = "lm", formula = y~x,show.legend = FALSE) +
  # stat_regline_equation(label.x = 2, #label.y = 32
  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
  #   formula = y~x
  # ) +
  scale_colour_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12", "#007f7f", "#1f78b4", "#99d8c9", "#542788","black", "#d73027", "#EC7063","pink1")) +
scale_fill_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12", "#007f7f", "#1f78b4", "#99d8c9", "#542788","black", "#d73027", "#EC7063","pink1"))#+
  # facet_wrap(~region,ncol=2,scales = "free")

P_individual <- P_all+ylim(0,0.8)+
  facet_wrap(~sampleID,ncol=4)+#ylim(0.15,NA)+
  ggpubr::stat_regline_equation(label.x = 0.2, label.y = 0.65,
    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,size=6,
    formula = y~x
  )+
  theme_cowplot(20)+
  theme(
    strip.background = element_rect(fill = "white", colour = NA),
    strip.text = element_text(colour = "black"),
    panel.border = element_rect(colour="black")
  )

P_all/P_individual+plot_layout(heights = c(0.8,0.2))

@@

# References

```{r Modern iso ranges}
# Ranges <- D18O %>%
#   filter(!(sampleID == "ORK_LT5" & d18O < 0.2)) %>% #excluding outlier
# group_by(sampleID,period) %>%
#   mutate(max=max(d18O),
#          min=min(d18O)) %>%
#   select(min, sampleID,max) %>% unique()
# 
# Modern_Ranges <- Ranges %>%
#   filter(period=="Modern") %>%
#   arrange(range)

@@

```{r Correlation Graphs, eval=FALSE }
#| fig-cap: Correlation graphs for all specimens
#| fig-height: 20
#| fig-width: 10

### 2024-06-24 Manuscript.tex
added separate graphs

@@

Figure~\ref{fig-Nac_iso} and Figure~\ref{fig-Pat_iso} show the
previously acquired sequential δ\textsuperscript{18}O-values of limpet
shells by species. Due to the high sampling resolution, all shells show
seasonally resolved and multi-year quasi-sinusoidal records of SST
change , with \emph{Nacella} sp. shells living 2--3 years and
\emph{Patella v}. regularly over 3 years. In those instances where
longer lifetimes are encountered (i.e.~specimen QG1-7246-1,
Figure~\ref{fig-Pat_iso} b). Interestingly, the ranges in
δ\textsuperscript{18}O-values are not similar between the coeval
\emph{Nacella} sp. specimens with δ\textsuperscript{18}O values for the
previous summer temperatures ranging from 0.4‰ to above 1.0‰, and for
winter values between 3.0‰ and 2.0‰. \emph{P. vulgata} has similarly
variable δ\textsuperscript{18}O-values for summer and winter seasons,
alas the lack of more that one modern specimens, means that we cannot
quantify the variability of recorded annual ranges between specimens.

@@

\caption{\label{fig-Nac_iso}Isotope values of Nacella sp. limpet shells
from Cambaceres Bay (\emph{a-f}).}

@@

\caption{\label{fig-Pat_iso}Isotope values of Patella vulagata limpet
shells from Quoygrew (QG, \emph{a-g}) and Rack Wick Bay (ORK, \emph{h})
on Westray, Orkney.}

@@

@@

\subsection{\texorpdfstring{\emph{Nacella
deaureata}}{Nacella deaureata}}\label{nacella-deaureata}

\subsection{\texorpdfstring{\emph{Nacella
magellanica}}{Nacella magellanica}}\label{nacella-magellanica}

\section{}\label{section}

@@

{\centering \includegraphics{Manuscript_files/figure-pdf/Correlation Graphs-1.pdf}

@@

\caption{Correlation graphs for all specimens}

>>>>

Using this system we carried out elemental imaging of the shell
specimens at a resolution of 50 µm distance between sampling spots. Each
spot was irradiated 10 times with the 3 first spectra used for cleaning
steps and the remaining summed to get an average Mg/Ca intensity ratio
for each sample spot. Subsequently, we re-sampled the section using a
line scan at 10 µm resolution. This leads to an overlap between sample
spots, but also allows for a continuous record without gaps. Intensity
ratios were filtered for cases with high relative standard deviation
(i.e.~more than 10\%). These occurred in places where the previous
sampling procedures for carbonate powder as part of the oxygen isotope
analysis left an uneven sample surfaces introducing variability in the
plasma generation and thus uncertainty in the data of these locations.

@@

Figure~\ref{fig-Nac_iso} and Figure~\ref{fig-Pat_iso} show the
previously acquired sequential δ\textsuperscript{18}O-values of limpet
shells by species. Due to the high sampling resolution, all shells show
seasonally resolved and multi-year quasi-sinusoidal records of SST
change , with \emph{Nacella} sp. shells living 2--3 years and
\emph{Patella v}. regularly over 3 years. In those instances where
longer lifetimes are encountered (i.e.~specimen QG1-7246-1,
Figure~\ref{fig-Pat_iso} b). Interestingly, the annual ranges in
δ\textsuperscript{18}O-values are not similar between the coeval
\emph{Nacella} sp. specimens with δ\textsuperscript{18}O values for the
previous summer temperatures ranging from 0.4‰ to above 1.0‰, and for
winter values between 2.7‰ and 2.0‰. \emph{P. vulgata} has similarly
variable δ\textsuperscript{18}O-values for summer and winter seasons,
alas the lack of more that one modern specimens, means that we cannot
quantify the variability of recorded annual ranges between specimens.

@@

\caption{\label{fig-Nac_iso}Isotope values of \emph{Nacella} sp. limpet
shells from Cambaceres Bay (\emph{a-f}).}

@@

\caption{\label{fig-Pat_iso}Isotope values of \emph{Patella vulgata}
limpet shells from Quoygrew (QG, \emph{a-g}) and Rack Wick Bay (ORK,
\emph{h}) on Westray, Orkney.}

@@

\subsection{Dynamic Time Warping}\label{dynamic-time-warping}

Because the exact location used in the analysis of stable oxygen
isotopes is not possible to reconstruct for each carbonate powder
sample, we used dynamic time warping (DTW) to align the time series of
δ\textsuperscript{18}O-values and Mg/Ca ratios. DTW is an algorithm that
measures similarity between two proxy-sequences, which may vary in
sampling resolution or interval. By stretching or compressing sections
of the series, DTW finds the probable alignment between the two
sequences. This allows us to compare the proxy data sets more
effectively, ensuring that the temporal dynamics of each shell are
accurately matched despite possible discrepancies in sampling intervals
or rates. We applied the DTW algorithm using the \texttt{dtw} package in
R \citep{Giorgino2009-sj, R_Core_Team2020-mk}, which provides a robust
framework for aligning time series data. This involved selecting
appropriate distance measures and constraints to ensure meaningful
alignment. The process involved iterative adjustments to minimise the
overall distance between corresponding points in the two data sets. The
code used for this analysis as well as the data required to run it can
be found in this \textbf{OSF repository}.

@@

\begin{itemize}
\tightlist
\item
  Show one or a few of the LIBS maps as well as the line scans.
\item
  Show one the line scans together with isotope data.
\item
  Show correlation graphs.
\end{itemize}

@@

{\centering \includegraphics{Manuscript_files/figure-pdf/Patella_Correlation_Graphs-1.pdf}

@@

\caption{Correlation graphs for \emph{Patella vulgata} specimens}

\subsection{\texorpdfstring{\emph{Nacella}
sp.}{Nacella sp.}}\label{nacella-sp.}

{\centering \includegraphics{Manuscript_files/figure-pdf/Nacella_Correlation_Graphs-1.pdf}

\caption{Correlation graphs for \emph{Nacella} sp. specimens}
